---
title: "Advanced R Workshop Instructor Guide: Plots and State Trends"
author: "BYU Library"
format: html
editor: visual
---

# Advanced R Workshop

This instructor guide uses the 50-state census-and-trends dataset. Each row is
a state, and the columns include climate, parks, UFO sightings, food, hobbies,
animals, and other trend scores.

## Instructor notes

All code blocks are complete and ready to run. The examples are designed to
demonstrate plot construction, faceting, and model output while reinforcing the
distinction between association and causation.

## Setup

Install these packages once if needed, then load them every time you start a
new R session.

```{r}
# Run once in the R console if you need the packages:
# install.packages(c("tidyverse", "broom"))

library(tidyverse)
library(broom)
```

Load the data. The `across()` step makes the four yes/no columns reliably
logical, and the new label columns make plots easier to read.

```{r}
census_df <- read.csv(
  "census_and_trends_data.csv",
  check.names = FALSE
) |>
  mutate(
    across(
      c(has_nba_team, has_major_zoo, is_coastal, has_waffle_house),
      ~ tolower(as.character(.x)) == "true"
    ),
    coastal_status = if_else(is_coastal, "Coastal", "Not coastal"),
    zoo_status = if_else(has_major_zoo, "Major zoo", "No major zoo"),
    waffle_house_status = if_else(
      has_waffle_house,
      "Has Waffle House",
      "No Waffle House"
    )
  )

glimpse(census_df)
```

# Plotting with ggplot2

`ggplot2` builds plots in layers. The usual pattern is:

```{r, eval = FALSE}
census_df |>
  ggplot(aes(x = x_variable, y = y_variable)) +
  geom_name()
```

The `aes()` function chooses what each visual property represents. The `geom_`
function chooses the kind of mark drawn on the graph.

## Scatterplot: `geom_point()`

This scatterplot shows annual sunshine and reported UFO sightings. Each point
represents one state.

```{r}
census_df |>
  ggplot(aes(x = annual_sunny_days, y = ufo_sightings_per_100k)) +
  geom_point(color = "darkorchid", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Annual sunshine and UFO sightings",
    subtitle = "A fitted trend line summarizes association, not causation.",
    x = "Annual sunny days",
    y = "UFO sightings per 100,000 people"
  ) +
  theme_minimal()
```

The black line is a fitted linear trend. It summarizes the pattern but does
not prove that sunshine causes, prevents, or inconveniences UFOs.

## Line plot: `geom_line()`

Line plots work best when the x-axis has a meaningful order. States do not
naturally occur in a numeric order, so first rank them from coldest to warmest.

```{r}
temperature_ranked <- census_df |>
  arrange(avg_temp_f) |>
  mutate(temperature_rank = row_number())

temperature_ranked |>
  ggplot(aes(x = temperature_rank, y = avg_temp_f)) +
  geom_line(color = "firebrick", linewidth = 1) +
  geom_point(color = "firebrick", size = 2) +
  labs(
    title = "Average temperature across states",
    subtitle = "States are ordered from coldest to warmest.",
    x = "Temperature rank",
    y = "Average temperature (°F)"
  ) +
  theme_minimal()
```

## Bar graph: `geom_bar()`

`geom_bar()` can draw bars from values you already calculated. First count the
number of states in each personality profile, then use `stat = "identity"` to
tell ggplot2 that the heights are already in `number_of_states`.

```{r}
profile_counts <- census_df |>
  count(personality_profile, name = "number_of_states") |>
  arrange(desc(number_of_states))

profile_counts |>
  ggplot(aes(x = reorder(personality_profile, number_of_states), y = number_of_states)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(
    title = "How the dataset labels state personality profiles",
    x = NULL,
    y = "Number of states"
  ) +
  theme_minimal()
```

## Histogram: `geom_histogram()`

A histogram shows the distribution of one numeric variable. Here we ask how
common different levels of UFO sightings are.

```{r}
census_df |>
  ggplot(aes(x = ufo_sightings_per_100k)) +
  geom_histogram(binwidth = 10, fill = "mediumpurple", color = "white") +
  labs(
    title = "The distribution of reported UFO sightings",
    x = "UFO sightings per 100,000 people",
    y = "Number of states"
  ) +
  theme_minimal()
```

## Box plot: `geom_boxplot()`

A box plot compares distributions across groups. This one compares average
temperature across personality profiles.

```{r}
census_df |>
  ggplot(aes(x = personality_profile, y = avg_temp_f)) +
  geom_boxplot(fill = "gold", alpha = 0.8) +
  coord_flip() +
  labs(
    title = "Average temperature by personality profile",
    x = NULL,
    y = "Average temperature (°F)"
  ) +
  theme_minimal()
```

# Faceting plots

Faceting creates a small plot for each subgroup. It is especially useful when
one busy plot would hide important differences between groups.

## `facet_wrap()`

Use `facet_wrap()` to make one panel for each personality profile. This lets
us compare the sunshine-and-UFO pattern within each group.

```{r}
census_df |>
  ggplot(aes(x = annual_sunny_days, y = ufo_sightings_per_100k)) +
  geom_point(color = "darkorchid", alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  facet_wrap(~ personality_profile) +
  labs(
    title = "Sunshine and UFO sightings by personality profile",
    x = "Annual sunny days",
    y = "UFO sightings per 100,000 people"
  ) +
  theme_minimal()
```

## `facet_grid()`

Use `facet_grid()` when two variables define the panels. This histogram breaks
average temperature into four combinations: coastal/not coastal and major
zoo/no major zoo.

```{r}
census_df |>
  ggplot(aes(x = avg_temp_f)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "white") +
  facet_grid(coastal_status ~ zoo_status) +
  labs(
    title = "Average temperature by coast and zoo status",
    x = "Average temperature (°F)",
    y = "Number of states"
  ) +
  theme_minimal()
```

When comparing facets, check whether axes are shared. Keeping the same axes
makes visual comparisons fair; changing them can make small differences look
dramatic.

# Statistical functions

Statistical tests can help quantify a pattern, but they do not automatically
make a causal claim. The `broom` package turns many model and test results into
clean data frames that work nicely with the `tidyverse`.

## T-test: `t.test()`

Do coastal states have a different average temperature than non-coastal states?
The t-test compares the means of two groups.

```{r}
coastal_temperature_test <- t.test(
  avg_temp_f ~ coastal_status,
  data = census_df
)

coastal_temperature_test
```

Use `tidy()` to turn the test result into a clean one-row data frame. It shows
the estimated difference, confidence interval, test statistic, and p-value.

```{r}
tidy(coastal_temperature_test)
```

Use `glance()` for a compact one-line summary of the same test.

```{r}
glance(coastal_temperature_test)
```

## Correlation: `cor()`

`cor()` measures the strength and direction of a linear relationship, from
`-1` to `1`. A value near zero indicates little linear relationship.

```{r}
cor(
  census_df$annual_sunny_days,
  census_df$ufo_sightings_per_100k
)
```

The result is a modest negative correlation in this dataset. That does **not**
mean clouds summon UFOs. It only describes how the two variables move together
in these 50 rows.

For a second suspiciously specific question, test whether states that search
more for capybaras also search more for Dungeons & Dragons.

```{r}
cor(
  census_df$capybara,
  census_df$`dungeons and dragons`
)
```

### A correlation sampler

Try several different kinds of pairings before deciding whether any pattern is
interesting, plausible, or merely hilarious. This code calculates five
correlations and sorts them by absolute size.

```{r}
correlation_sampler <- data.frame(
  pair = c(
    "Waffles and pancakes",
    "National parks and stargazing",
    "Dog ownership and hiking",
    "Craft breweries and board games",
    "Video games and Dungeons & Dragons"
  ),
  correlation = c(
    cor(census_df$waffles, census_df$pancakes),
    cor(census_df$national_parks_count, census_df$stargazing),
    cor(census_df$dog_ownership_pct, census_df$hiking),
    cor(census_df$craft_breweries_per_100k, census_df$`board games`),
    cor(census_df$`video games`, census_df$`dungeons and dragons`)
  )
) |>
  arrange(desc(abs(correlation)))

correlation_sampler
```

The strongest result in a short list is not automatically important. It may be
a real relationship, an accident of a small sample, a proxy for another factor,
or a clue that the trend data deserves a closer look. In particular, do not
conclude that playing video games prevents people from rolling a d20, or that
board games cause new breweries to appear.

Here is a scatterplot for the video-games-and-Dungeons-&-Dragons pairing. A
plot helps you see whether a correlation comes from a broad pattern or a few
unusual states.

```{r}
census_df |>
  ggplot(aes(x = `video games`, y = `dungeons and dragons`)) +
  geom_point(color = "seagreen4", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Video game and Dungeons & Dragons trend scores",
    subtitle = "A relationship is not a cultural explanation.",
    x = "Video games trend score",
    y = "Dungeons & Dragons trend score"
  ) +
  theme_minimal()
```

### Short demonstrations

These examples draw from different parts of the dataset and are useful for
brief live demonstrations of group comparisons and correlation plots.

#### NBA teams and taco trend scores

This summary compares mean taco trend scores for states with and without an NBA
team.

```{r}
nba_taco_summary <- census_df |>
  group_by(has_nba_team) |>
  summarise(
    average_taco_score = mean(tacos),
    number_of_states = n(),
    .groups = "drop"
  ) |>
  mutate(
    nba_team_status = if_else(
      has_nba_team,
      "Has NBA team",
      "No NBA team"
    )
  )

nba_taco_summary
```

```{r}
nba_taco_summary |>
  ggplot(aes(x = nba_team_status, y = average_taco_score)) +
  geom_bar(stat = "identity", fill = "tomato") +
  labs(
    title = "Average taco trend score by NBA-team status",
    subtitle = "Basketball and tacos may share an arena. Or they may not.",
    x = NULL,
    y = "Average taco trend score"
  ) +
  theme_minimal()
```

The t-test quantifies the observed group difference.

```{r}
nba_taco_test <- t.test(tacos ~ has_nba_team, data = census_df)

tidy(nba_taco_test)
```

#### State population and capybara trend scores

Population has a very large range, so a log-scale x-axis makes the points
easier to see.

```{r}
cor(census_df$total_pop, census_df$capybara)

census_df |>
  ggplot(aes(x = total_pop, y = capybara)) +
  geom_point(color = "darkorange", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  scale_x_log10() +
  labs(
    title = "State population and capybara trend scores",
    subtitle = "A larger population may mean more capybara enthusiasts—or just more people searching.",
    x = "Total population (log scale)",
    y = "Capybara trend score"
  ) +
  theme_minimal()
```

#### Dog ownership and hiking trend scores

This plot illustrates a correlation between dog ownership and hiking trend
scores. Interpret it as an association only.

```{r}
cor(census_df$dog_ownership_pct, census_df$hiking)

census_df |>
  ggplot(aes(x = dog_ownership_pct, y = hiking)) +
  geom_point(color = "forestgreen", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Dog ownership and hiking trend scores",
    x = "Dog ownership (%)",
    y = "Hiking trend score"
  ) +
  theme_minimal()
```

`cor.test()` adds a hypothesis test and confidence interval. `broom` can clean
up that result too.

```{r}
sunshine_ufo_test <- cor.test(
  census_df$annual_sunny_days,
  census_df$ufo_sightings_per_100k
)

tidy(sunshine_ufo_test)
glance(sunshine_ufo_test)
```

## Regression: `lm()`

Linear regression models a numeric outcome from one or more predictors. This
model estimates UFO sightings from annual sunshine and average temperature.

```{r}
ufo_model <- lm(
  ufo_sightings_per_100k ~ annual_sunny_days + avg_temp_f,
  data = census_df
)

ufo_model
```

Use `tidy()` to see one row per coefficient and `glance()` to see model-level
information such as (R^2).

```{r}
tidy(ufo_model)
glance(ufo_model)
```

Use `augment()` to put fitted values and residuals back beside the original
data. A residual is the observed value minus the model's predicted value.

```{r}
ufo_predictions <- augment(ufo_model, data = census_df)

ufo_predictions |>
  select(state, ufo_sightings_per_100k, .fitted, .resid) |>
  arrange(desc(abs(.resid))) |>
  head(10)
```

The states at the top are the ones whose UFO-sighting rates differ the most
from what this simple model predicts. They are unusual for this model, not
proof of alien activity.

## ANOVA: `aov()`

ANOVA compares the mean of a numeric variable across three or more groups.
This analysis compares average temperatures across the three personality
profiles.

```{r}
temperature_anova <- aov(
  avg_temp_f ~ personality_profile,
  data = census_df
)

summary(temperature_anova)
```

Use `tidy()` to turn the ANOVA table into a clean data frame, then `glance()`
for a compact model summary.

```{r}
tidy(temperature_anova)
glance(temperature_anova)
```

# Practice challenges

Try these with the same data:

1. Make a scatterplot of `waffles` versus `pancakes`. Add a linear trend line.
2. Make a histogram of `dog_ownership_pct` with a bin width of 5.
3. Make a box plot of `hiking` scores by `coastal_status`.
4. Use `facet_wrap()` to split a pizza-versus-hiking scatterplot by
   `personality_profile`.
5. Run a t-test comparing `dog_ownership_pct` for states with and without a
   Waffle House. Then clean the result with `tidy()`.
6. Fit a regression predicting `capybara` trend scores from `waffles` and
   `dungeons and dragons`. Use `augment()` to find the largest residuals.

## Find your own craziest correlation

Choose any two numeric columns in the dataset that make you curious. Calculate
their correlation, make a scatterplot, and then inspect the points for unusual
states or outliers. Finish by writing one sentence about what the pattern does
**not** prove and one possible third factor that could influence both variables.

## Instructor answer key

### Challenge 1: waffles and pancakes

```{r}
census_df |>
  ggplot(aes(x = waffles, y = pancakes)) +
  geom_point(color = "goldenrod", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Waffle and pancake trend scores",
    x = "Waffle trend score",
    y = "Pancake trend score"
  ) +
  theme_minimal()
```

### Challenge 2: dog ownership histogram

```{r}
census_df |>
  ggplot(aes(x = dog_ownership_pct)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of dog ownership across states",
    x = "Dog ownership (%)",
    y = "Number of states"
  ) +
  theme_minimal()
```

### Challenge 3: hiking by coastal status

```{r}
census_df |>
  ggplot(aes(x = coastal_status, y = hiking)) +
  geom_boxplot(fill = "seagreen3", alpha = 0.8) +
  labs(
    title = "Hiking trend scores by coastal status",
    x = NULL,
    y = "Hiking trend score"
  ) +
  theme_minimal()
```

### Challenge 4: pizza and hiking by personality profile

```{r}
census_df |>
  ggplot(aes(x = pizza, y = hiking)) +
  geom_point(color = "firebrick", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  facet_wrap(~ personality_profile) +
  labs(
    title = "Pizza and hiking trend scores by personality profile",
    x = "Pizza trend score",
    y = "Hiking trend score"
  ) +
  theme_minimal()
```

### Challenge 5: dog ownership and Waffle House status

```{r}
dog_waffle_test <- t.test(
  dog_ownership_pct ~ waffle_house_status,
  data = census_df
)

tidy(dog_waffle_test)
```

### Challenge 6: capybara regression

```{r}
capybara_model <- lm(
  capybara ~ waffles + `dungeons and dragons`,
  data = census_df
)

capybara_model |>
  augment(data = census_df) |>
  select(state, capybara, .fitted, .resid) |>
  arrange(desc(abs(.resid))) |>
  head(10)
```

# Teaching close

Close by reinforcing that a compelling plot, correlation, or p-value is a
starting point for analysis. Causal claims require a research design that can
rule out alternative explanations.
