---
title: "Advanced R Workshop Instructor Guide: Ice Cream and State Trends"
author: "BYU Library"
format: html
editor: visual
---

# Advanced R Workshop

This instructor guide uses the 50-state census-and-trends dataset. Each row is
a state, and the columns include climate, parks, UFO sightings, food, hobbies,
animals, and other trend scores. Our running question follows beginning R:
**Do colder states have higher ice cream trend scores?**

## Instructor notes

All code blocks are complete and ready to run. The examples are designed to
demonstrate plot construction, faceting, and model output while reinforcing the
distinction between association and causation.

`ice cream` is a trend score, not a measure of ice cream sales or
consumption. The main question is a teaching example, not evidence that
weather changes what people eat.

## Setup

Install these packages once if needed, then load them every time you start a
new R session.

```{r}
# Run once in the R console if you need the packages:
# install.packages(c("tidyverse", "broom"))

library(tidyverse)
library(broom)
```

Place the clean October 7 CSV beside this file, then load it. The `across()` step makes the four yes/no columns reliably
logical, and the new label columns make plots easier to read.

```{r}
census_df <- read.csv(
  "census_and_trends_data_2026-10-7.csv",
  check.names = FALSE
) |>
  mutate(
    across(
      c(has_nba_team, has_major_zoo, is_coastal, has_waffle_house),
      ~ tolower(as.character(.x)) == "true"
    ),
    coastal_status = if_else(is_coastal, "Coastal", "Not coastal"),
    temperature_group = if_else(
      avg_temp_f < 50,
      "Below 50°F",
      "50°F or warmer"
    ),
    zoo_status = if_else(has_major_zoo, "Major zoo", "No major zoo"),
    waffle_house_status = if_else(
      has_waffle_house,
      "Has Waffle House",
      "No Waffle House"
    )
  )

glimpse(census_df)
```

# Plotting with ggplot2

`ggplot2` builds plots in layers. The usual pattern is:

```{r, eval = FALSE}
census_df |>
  ggplot(aes(x = x_variable, y = y_variable)) +
  geom_name()
```

The `aes()` function chooses what each visual property represents. The `geom_`
function chooses the kind of mark drawn on the graph.

## Scatterplot: `geom_point()`

Start with the plot from beginning R, then add a fitted line. Each point
represents one state. Do colder states have higher ice cream trend scores?

```{r}
census_df |>
  ggplot(aes(x = avg_temp_f, y = `ice cream`)) +
  geom_point(color = "deepskyblue4", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Do colder states have higher ice cream trend scores?",
    subtitle = "A trend line describes association, not ice cream consumption.",
    x = "Average temperature (°F)",
    y = "Ice cream trend score"
  ) +
  theme_minimal()
```

The black line summarizes a pattern across the states. It does not establish
that temperatures caused the difference in trend scores.

## Line plot: `geom_line()`

Line plots work best when the x-axis has a meaningful order. States do not
naturally occur in a numeric order, so first rank them from coldest to warmest.

```{r}
temperature_ranked <- census_df |>
  arrange(avg_temp_f) |>
  mutate(temperature_rank = row_number())

temperature_ranked |>
  ggplot(aes(x = temperature_rank, y = avg_temp_f)) +
  geom_line(color = "firebrick", linewidth = 1) +
  geom_point(color = "firebrick", size = 2) +
  labs(
    title = "Average temperature across states",
    subtitle = "States are ordered from coldest to warmest.",
    x = "Temperature rank",
    y = "Average temperature (°F)"
  ) +
  theme_minimal()
```

## Bar graph: `geom_bar()`

`geom_bar()` can draw bars from values you already calculated. First count the
number of states in each personality profile, then use `stat = "identity"` to
tell ggplot2 that the heights are already in `number_of_states`.

```{r}
profile_counts <- census_df |>
  count(personality_profile, name = "number_of_states") |>
  arrange(desc(number_of_states))

profile_counts |>
  ggplot(aes(x = reorder(personality_profile, number_of_states), y = number_of_states)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(
    title = "How the dataset labels state personality profiles",
    x = NULL,
    y = "Number of states"
  ) +
  theme_minimal()
```

## Histogram: `geom_histogram()`

A histogram shows the distribution of one numeric variable. Here we ask how
common different ice cream trend scores are.

```{r}
census_df |>
  ggplot(aes(x = `ice cream`)) +
  geom_histogram(binwidth = 3, fill = "mediumpurple", color = "white") +
  labs(
    title = "How do ice cream trend scores vary across states?",
    x = "Ice cream trend score",
    y = "Number of states"
  ) +
  theme_minimal()
```

## Box plot: `geom_boxplot()`

A box plot compares distributions across groups. This one compares ice cream
trend scores for states below and above the 50°F cutoff from beginning R.

```{r}
census_df |>
  ggplot(aes(x = temperature_group, y = `ice cream`)) +
  geom_boxplot(fill = "gold", alpha = 0.8) +
  coord_flip() +
  labs(
    title = "Ice cream trend scores in colder and warmer states",
    x = NULL,
    y = "Ice cream trend score"
  ) +
  theme_minimal()
```

# Faceting plots

Faceting creates a small plot for each subgroup. It is especially useful when
one busy plot would hide important differences between groups.

## `facet_wrap()`

Use `facet_wrap()` to make one panel for each personality profile. This lets
us check whether the ice-cream-and-temperature pattern looks similar in all
three groups. There are only 13 states in the smallest profile, so
interpret its fitted line cautiously.

```{r}
census_df |>
  ggplot(aes(x = avg_temp_f, y = `ice cream`)) +
  geom_point(color = "deepskyblue4", alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  facet_wrap(~ personality_profile) +
  labs(
    title = "Temperature and ice cream trends by personality profile",
    x = "Average temperature (°F)",
    y = "Ice cream trend score"
  ) +
  theme_minimal()
```

## `facet_grid()`

Use `facet_grid()` when two variables define the panels. This histogram breaks
ice cream scores into four combinations: coastal/not coastal and major
zoo/no major zoo.

```{r}
census_df |>
  ggplot(aes(x = `ice cream`)) +
  geom_histogram(binwidth = 3, fill = "skyblue", color = "white") +
  facet_grid(coastal_status ~ zoo_status) +
  labs(
    title = "Ice cream scores by coast and zoo status",
    x = "Ice cream trend score",
    y = "Number of states"
  ) +
  theme_minimal()
```

When comparing facets, check whether axes are shared. Keeping the same axes
makes visual comparisons fair; changing them can make small differences look
dramatic.

# Statistical functions

Statistical tests can help quantify a pattern, but they do not automatically
make a causal claim. The `broom` package turns many model and test results into
clean data frames that work nicely with the `tidyverse`.

## T-test: `t.test()`

Do states below 50°F have a different mean ice cream trend score from warmer
states? The t-test compares the means of two groups. This cutoff was chosen
for the workshop and can be changed; it is not a natural divide.

```{r}
ice_cream_temperature_test <- t.test(
  `ice cream` ~ temperature_group,
  data = census_df
)

ice_cream_temperature_test
```

Use `tidy()` to turn the test result into a clean one-row data frame. It shows
the estimated difference, confidence interval, test statistic, and p-value.

```{r}
tidy(ice_cream_temperature_test)
```

Use `glance()` for a compact one-line summary of the same test.

```{r}
glance(ice_cream_temperature_test)
```

## Correlation: `cor()`

`cor()` measures the strength and direction of a linear relationship, from
`-1` to `1`. A value near zero indicates little linear relationship.

```{r}
cor(census_df$avg_temp_f, census_df$`ice cream`)
```

The result is about -0.12 in this version of the dataset: the linear
relationship between temperature and ice cream interest is weak. This does
**not** tell us how much ice cream residents eat, or why the scores differ.

For a second suspiciously specific question, test whether states that search
more for capybaras also search more for Dungeons & Dragons.

```{r}
cor(
  census_df$capybara,
  census_df$`dungeons and dragons`
)
```

### A correlation sampler

Try several different kinds of pairings before deciding whether any pattern is
interesting, plausible, or merely hilarious. This code calculates five
correlations and sorts them by absolute size.

```{r}
correlation_sampler <- data.frame(
  pair = c(
    "Waffles and pancakes",
    "National parks and stargazing",
    "Dog ownership and hiking",
    "Pizza and board games",
    "Video games and Dungeons & Dragons"
  ),
  correlation = c(
    cor(census_df$waffles, census_df$pancakes),
    cor(census_df$national_parks_count, census_df$stargazing),
    cor(census_df$dog_ownership_pct, census_df$hiking),
    cor(census_df$pizza, census_df$`board games`),
    cor(census_df$`video games`, census_df$`dungeons and dragons`)
  )
) |>
  arrange(desc(abs(correlation)))

correlation_sampler
```

The strongest result in a short list is not automatically important. It may be
a real relationship, an accident of a small sample, a proxy for another factor,
or a clue that the trend data deserves a closer look. In particular, do not
conclude that playing video games prevents people from rolling a d20, or that pizza searches cause game night.

### Optional data-quality detour: UFO sightings

The updated CSV gives each state a different UFO sighting value. Bin the values
to see the broader distribution before drawing a story from a single number:

```{r}
census_df |>
  mutate(ufo_band = cut(
    ufo_sightings_per_100k,
    breaks = seq(0, 100, by = 10),
    include.lowest = TRUE
  )) |>
  count(ufo_band, sort = TRUE)
```

Here is a scatterplot for the video-games-and-Dungeons-&-Dragons pairing. A
plot helps you see whether a correlation comes from a broad pattern or a few
unusual states.

```{r}
census_df |>
  ggplot(aes(x = `video games`, y = `dungeons and dragons`)) +
  geom_point(color = "seagreen4", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Video game and Dungeons & Dragons trend scores",
    subtitle = "A relationship is not a cultural explanation.",
    x = "Video games trend score",
    y = "Dungeons & Dragons trend score"
  ) +
  theme_minimal()
```

### Short demonstrations

These examples draw from different parts of the dataset and are useful for
brief live demonstrations of group comparisons and correlation plots.

#### NBA teams and taco trend scores

This summary compares mean taco trend scores for states with and without an NBA
team.

```{r}
nba_taco_summary <- census_df |>
  group_by(has_nba_team) |>
  summarise(
    average_taco_score = mean(tacos),
    number_of_states = n(),
    .groups = "drop"
  ) |>
  mutate(
    nba_team_status = if_else(
      has_nba_team,
      "Has NBA team",
      "No NBA team"
    )
  )

nba_taco_summary
```

```{r}
nba_taco_summary |>
  ggplot(aes(x = nba_team_status, y = average_taco_score)) +
  geom_bar(stat = "identity", fill = "tomato") +
  labs(
    title = "Average taco trend score by NBA-team status",
    subtitle = "Basketball and tacos may share an arena. Or they may not.",
    x = NULL,
    y = "Average taco trend score"
  ) +
  theme_minimal()
```

The t-test quantifies the observed group difference.

```{r}
nba_taco_test <- t.test(tacos ~ has_nba_team, data = census_df)

tidy(nba_taco_test)
```

#### State population and capybara trend scores

Population has a very large range, so a log-scale x-axis makes the points
easier to see.

```{r}
cor(census_df$total_pop, census_df$capybara)

census_df |>
  ggplot(aes(x = total_pop, y = capybara)) +
  geom_point(color = "darkorange", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  scale_x_log10() +
  labs(
    title = "State population and capybara trend scores",
    subtitle = "A larger population may mean more capybara enthusiasts—or just more people searching.",
    x = "Total population (log scale)",
    y = "Capybara trend score"
  ) +
  theme_minimal()
```

#### Dog ownership and hiking trend scores

This plot illustrates a correlation between dog ownership and hiking trend
scores. Interpret it as an association only.

```{r}
cor(census_df$dog_ownership_pct, census_df$hiking)

census_df |>
  ggplot(aes(x = dog_ownership_pct, y = hiking)) +
  geom_point(color = "forestgreen", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Dog ownership and hiking trend scores",
    x = "Dog ownership (%)",
    y = "Hiking trend score"
  ) +
  theme_minimal()
```

`cor.test()` adds a hypothesis test and confidence interval. `broom` can clean
up that result too.

```{r}
temperature_ice_cream_test <- cor.test(
  census_df$avg_temp_f,
  census_df$`ice cream`
)

tidy(temperature_ice_cream_test)
glance(temperature_ice_cream_test)
```

## Regression: `lm()`

Linear regression models a numeric outcome from one or more predictors. This
model estimates the ice cream trend score from average temperature and coastal
status. Compare its temperature coefficient with the direction of the plot.

```{r}
ice_cream_model <- lm(
  `ice cream` ~ avg_temp_f + coastal_status,
  data = census_df
)

ice_cream_model
```

Use `tidy()` to see one row per coefficient and `glance()` to see model-level
information such as (R^2).

```{r}
tidy(ice_cream_model)
glance(ice_cream_model)
```

Use `augment()` to put fitted values and residuals back beside the original
data. A residual is the observed value minus the model's predicted value.

```{r}
ice_cream_predictions <- augment(ice_cream_model, data = census_df)

ice_cream_predictions |>
  select(state, `ice cream`, avg_temp_f, coastal_status, .fitted, .resid) |>
  arrange(desc(abs(.resid))) |>
  head(10)
```

The states at the top are the ones whose ice cream trend scores differ the
most from this simple model's predictions. Ask which other variables might
help explain the surprise; a large residual is not evidence of a cause.

## ANOVA: `aov()`

ANOVA compares the mean of a numeric variable across three or more groups.
This analysis compares ice cream trend scores across the three personality
profiles. Compare this result with the weak temperature relationship. Different
groupings can reveal different patterns without proving a cause.

```{r}
ice_cream_anova <- aov(
  `ice cream` ~ personality_profile,
  data = census_df
)

summary(ice_cream_anova)
```

Use `tidy()` to turn the ANOVA table into a clean data frame, then `glance()`
for a compact model summary.

```{r}
tidy(ice_cream_anova)
glance(ice_cream_anova)
```

# Practice challenges

Try these with the same data:

1. Make a scatterplot of `waffles` versus `pancakes`. Add a linear trend line.
2. Make a histogram of `dog_ownership_pct` with a bin width of 5.
3. Make a box plot of `hiking` scores by `coastal_status`.
4. Use `facet_wrap()` to split a pizza-versus-hiking scatterplot by
   `personality_profile`.
5. Run a t-test comparing `dog_ownership_pct` for states with and without a
   Waffle House. Then clean the result with `tidy()`.
6. Fit a regression predicting `capybara` trend scores from `waffles` and
   `dungeons and dragons`. Use `augment()` to find the largest residuals.
7. **Cold scoop detective:** Facet the ice-cream-versus-temperature plot by
   `coastal_status`. Add a trend line in each panel. Then use the fitted
   `ice_cream_model` to identify the state with the biggest *positive*
   residual: which state is more ice-cream-curious than the model expects?
   Write a headline that says "trend score," not "ice cream sales."
8. **Alien sourdough hypothesis (optional):** Do UFO reports rise in states
   with higher sourdough-baking search interest? Calculate a correlation and
   make a scatterplot. Then write a wonderfully dramatic headline that does
   **not** claim sourdough attracts UFOs.

## Find your own craziest correlation

Choose any two numeric columns in the dataset that make you curious. Calculate
their correlation, make a scatterplot, and then inspect the points for unusual
states or outliers. Finish by writing one sentence about what the pattern does
**not** prove and one possible third factor that could influence both variables.

## Instructor answer key

### Challenge 1: waffles and pancakes

```{r}
census_df |>
  ggplot(aes(x = waffles, y = pancakes)) +
  geom_point(color = "goldenrod", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Waffle and pancake trend scores",
    x = "Waffle trend score",
    y = "Pancake trend score"
  ) +
  theme_minimal()
```

### Challenge 2: dog ownership histogram

```{r}
census_df |>
  ggplot(aes(x = dog_ownership_pct)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of dog ownership across states",
    x = "Dog ownership (%)",
    y = "Number of states"
  ) +
  theme_minimal()
```

### Challenge 3: hiking by coastal status

```{r}
census_df |>
  ggplot(aes(x = coastal_status, y = hiking)) +
  geom_boxplot(fill = "seagreen3", alpha = 0.8) +
  labs(
    title = "Hiking trend scores by coastal status",
    x = NULL,
    y = "Hiking trend score"
  ) +
  theme_minimal()
```

### Challenge 4: pizza and hiking by personality profile

```{r}
census_df |>
  ggplot(aes(x = pizza, y = hiking)) +
  geom_point(color = "firebrick", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  facet_wrap(~ personality_profile) +
  labs(
    title = "Pizza and hiking trend scores by personality profile",
    x = "Pizza trend score",
    y = "Hiking trend score"
  ) +
  theme_minimal()
```

### Challenge 5: dog ownership and Waffle House status

```{r}
dog_waffle_test <- t.test(
  dog_ownership_pct ~ waffle_house_status,
  data = census_df
)

tidy(dog_waffle_test)
```

### Challenge 6: capybara regression

```{r}
capybara_model <- lm(
  capybara ~ waffles + `dungeons and dragons`,
  data = census_df
)

capybara_model |>
  augment(data = census_df) |>
  select(state, capybara, .fitted, .resid) |>
  arrange(desc(abs(.resid))) |>
  head(10)
```

### Challenge 7: cold scoop detective

```{r}
census_df |>
  ggplot(aes(x = avg_temp_f, y = `ice cream`)) +
  geom_point(color = "deepskyblue4", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  facet_wrap(~ coastal_status) +
  labs(
    title = "Ice cream trend scores and temperature by coast",
    x = "Average temperature (°F)",
    y = "Ice cream trend score"
  ) +
  theme_minimal()

ice_cream_predictions |>
  select(state, avg_temp_f, `ice cream`, .fitted, .resid) |>
  arrange(desc(.resid)) |>
  head(1)
```

### Challenge 8: alien sourdough hypothesis

```{r}
cor(census_df$ufo_sightings_per_100k, census_df$`sourdough baking`)

census_df |>
  ggplot(aes(x = `sourdough baking`, y = ufo_sightings_per_100k)) +
  geom_point(color = "darkorchid4", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "UFO reports and sourdough-baking searches",
    x = "Sourdough-baking search-interest score",
    y = "UFO sightings per 100,000 people"
  ) +
  theme_minimal()
```

The correlation is about 0.70 in this dataset. It cannot tell us whether
either variable influences the other, or how these trend scores were produced.

# Teaching close

Close by reinforcing that a compelling plot, correlation, or p-value is a
starting point for analysis. Causal claims require a research design that can
rule out alternative explanations.

Because these 50 rows are states rather than a random sample of people,
statistical p-values are examples of test output, not proof of a general
claim about all ice cream lovers. Trend scores are not sales or consumption.
