---
title: "R Workshop: Census and Trends"
author: "BYU Library"
format: html
editor: visual
---

# Basics of R

## Your first five minutes

This is an instructor guide with complete solutions. For the live workshop,
click the run button beside one `{r}` code block at a time and look at its
result before moving on. If a block produces an error, pause there and read
the message together. A `#` line is a note for humans; R does not run it.

**Live route (about one hour):** Try one calculator change, load the data,
practice `select()` and `filter()`, sort the cold ice cream states, make one
plot, and finish with the cold scoop challenge. The other examples stay in
place for quick demonstrations or later practice.

## What is R?

R is a programming language that can act like a calculator and can also perform more complex data-analysis tasks. For this workshop, we will focus on three general types: *numbers*, *characters*, and *variables*. Our running question is: **Do colder states have higher ice cream trend scores?**

```{r}
# Numbers
6
20.5
13 + 30

# Characters
"waffles"
"Brigham Young University"

# Variables
my_favorite_number <- 6
road_trip_snack <- "waffles"

# Try it: replace 6 and "waffles" with your own choices, then run again.
my_favorite_number
road_trip_snack
```

R uses **functions** to manipulate and analyze data. Functions use parentheses:

```{r}
# Prints text in quotes
print("Hello, data explorer!")

# Takes the square root of a number
sqrt(36)
```

**Libraries**, or **packages**, are toolboxes of functions written by other people. `tidyverse`
includes `dplyr` for working with tables and `ggplot2` for plotting. Install
it once if needed; load it each time you start R.

```{r}
# Run once in the R console only if tidyverse is not installed:
# install.packages("tidyverse")
library(tidyverse)
```

## Key Concepts

The assignment operator is different from the **comparison operator** (`==`). A comparison returns either `TRUE` or `FALSE`.

```{r}
5 == 5
"waffles" == "tacos"
```

We use the **assignment operator** (`<-` or `=`) to store information in variables. It means "gets saved as."

```{r}
# The beginning R workshop uses its own link to the clean CSV.
census_df <- read_csv("https://hbll.s.gy/rbegin")
```

A **data frame** is a table or spreadsheet. Each row is an observation, and each column is a variable. In this workshop, each row is a U.S. state, with census-style facts and trend scores for food, sports, animals, and hobbies.

```{r}
# Start with only four columns so the wide data table is readable.
census_df |>
  select(state, avg_temp_f, `ice cream`, has_major_zoo) |>
  head(6)
```

# Census and Trends Data Frame Analysis

## Analyzing the Data

During this workshop, we will use these common `dplyr` functions:

- `select()` keeps chosen columns
- `filter()` keeps rows that meet a condition
- `mutate()` creates or changes columns
- `group_by()` creates groups for a later calculation
- `summarise()` calculates a summary for each group
- `glimpse()` previews a data frame's structure
- `arrange()` sorts rows

### Selecting columns

Read `|>` as “then”: start with the data, *then* choose columns with
`select()`. Names with spaces, like `ice cream`, need backticks in R code.
**Live:** Run the example once, then replace `has_major_zoo` with
`annual_sunny_days` and run it again. Which column changed?

Suppose you are planning an ice cream road trip. Select the state name, average temperature, ice cream trend score, and major-zoo status.

```{r}
census_df |>
  select(
    state,
    avg_temp_f,
    `ice cream`,
    has_major_zoo
  )
```

### Filtering rows

To see states whose average temperature is below 50°F:
**Live:** Read `avg_temp_f < 50` as “temperature is below 50.” Run the code,
then try `avg_temp_f < 45` to see fewer states. Restore 50 afterward.

```{r}
census_df |>
  filter(avg_temp_f < 50)
```

### Creating columns

**Instructor demonstration:** Show what `mutate()` produces. Learners can
return to the brunch score during practice.

The dataset includes trend scores for waffles, pancakes, and donuts. Create a simple `brunch_score` by adding them together.

```{r}
census_df |>
  mutate(brunch_score = waffles + pancakes + donuts)
```

`mutate()` displays a new data frame. To save the new column for later, assign the result back to a variable.

```{r}
census_with_brunch <- census_df |>
  mutate(brunch_score = waffles + pancakes + donuts)
```

### Grouping and summarizing

**Instructor demonstration or later practice:** It is enough today to notice
that `group_by()` makes groups and `summarise()` gives one answer per group.

The states are grouped into personality profiles. Find the average temperature, average annual sunshine, average UFO sightings, and number of states in each profile.

```{r}
census_df |>
  group_by(personality_profile) |>
  summarise(
    average_temperature_f = mean(avg_temp_f),
    average_sunny_days = mean(annual_sunny_days),
    average_ufo_sightings = mean(ufo_sightings_per_100k),
    number_of_states = n(),
    .groups = "drop"
  )
```

### Inspecting a data frame

Use `glimpse()` to see column names, data types, and a compact preview of the data. Selecting a few columns first keeps the output easy to read.

```{r}
census_df |>
  select(
    state,
    personality_profile,
    avg_temp_f,
    annual_sunny_days,
    ufo_sightings_per_100k
  ) |>
  glimpse()
```

### Sorting rows

**Live:** `desc()` means descending, or highest first. Before running this
block, predict one sunny state that might appear near the top.

Use `arrange()` to put rows in order. Here are the ten sunniest states, from most to least sunny:

```{r}
census_df |>
  arrange(desc(annual_sunny_days)) |>
  select(state, annual_sunny_days) |>
  head(10)
```

That is enough to begin exploring the data!

## Ice cream mystery: do colder states score higher?

`ice cream` is a trend score in this dataset. It does not measure how much ice cream residents buy or eat. First, find the cold states with the highest ice cream scores:

**Live:** This repeats familiar moves: `filter()` keeps cold states,
`arrange()` puts higher scores first, and `select()` shows three columns.
Read each line as “then” before you run the whole block.

```{r}
census_df |>
  filter(avg_temp_f < 50) |>
  arrange(desc(`ice cream`)) |>
  select(state, avg_temp_f, `ice cream`) |>
  head(10)
```

Compare the average scores for colder and warmer states. The 50°F cutoff is our choice for this activity, not a scientific definition of a cold state.

**Instructor demonstration:** Explain the two resulting rows without
requiring learners to write `mutate()`, `group_by()`, and `summarise()` today.

```{r}
census_df |>
  mutate(temperature_group = if_else(avg_temp_f < 50, "Below 50°F", "50°F or warmer")) |>
  group_by(temperature_group) |>
  summarise(
    states = n(),
    average_ice_cream_score = mean(`ice cream`),
    .groups = "drop"
  )
```

A simple scatterplot helps us see whether the pattern holds across states. Advanced R will return to this plot to add a trend line, facets, and a model.

**Live:** Run the plot, then change the title in the `title =` line to your own
question. In ggplot code, `+` adds a piece to the plot; `|>` still means “then.”

```{r}
census_df |>
  ggplot(aes(x = avg_temp_f, y = `ice cream`)) +
  geom_point(color = "deepskyblue4", size = 2.5) +
  labs(
    title = "Do colder states have higher ice cream trend scores?",
    x = "Average temperature (°F)",
    y = "Ice cream trend score"
  ) +
  theme_minimal()
```

## Practice Time

**Live closing challenge:** Start by changing one condition in the cold-state
filter you already ran: add `has_major_zoo` as another condition, then reuse
the `arrange()` and `select()` lines. The complete instructor solution is
below. Use the remaining exercises for early finishers or later practice.

**Cold scoop challenge:** Predict a state first! Your friend wants to visit a major zoo and find a state with a high ice cream trend score. Among states below 50°F that have a major zoo, which state ranks highest? What happens if you remove the zoo requirement? A trend score does not confirm what a zoo serves.

```{r}
# INSTRUCTOR SOLUTION: let participants try first
census_df |>
  filter(avg_temp_f < 50, has_major_zoo) |>
  arrange(desc(`ice cream`)) |>
  select(state, avg_temp_f, `ice cream`, has_major_zoo) |>
  head(5)
```

**Filtering:** Find states with at least 250 sunny days per year. These might be good candidates for a sunny road trip.

```{r}
# COMPLETED CODE BELOW
census_df |>
  filter(annual_sunny_days >= 250)
```

Find the row for Utah. Does its sunshine suit your next road-trip stop?

```{r}
# COMPLETED CODE BELOW
census_df |>
  filter(state == "Utah")
```

**Mutating:** Create a `brunch_score` column that adds the trend scores for waffles, pancakes, and donuts.

```{r}
# COMPLETED CODE BELOW
census_df |>
  mutate(brunch_score = waffles + pancakes + donuts)
```

Create a Boolean `UFO_hotspot` column that is `TRUE` when a state has at least 60 UFO sightings per 100,000 people.

```{r}
# COMPLETED CODE BELOW
census_df |>
  mutate(UFO_hotspot = ufo_sightings_per_100k >= 60)
```

**Selecting:** Create a smaller data frame with a state's personality profile, capybara trend score, and Dungeons & Dragons trend score.

```{r}
# COMPLETED CODE BELOW
census_df |>
  select(state, personality_profile, capybara, `dungeons and dragons`)
```

Now select only the state names.

```{r}
# COMPLETED CODE BELOW
census_df |>
  select(state)
```

**Summarizing:** Find the total population represented in the dataset.

```{r}
# COMPLETED CODE BELOW
census_df |>
  summarise(total_population = sum(total_pop))
```

Find the coldest average temperature in the dataset.

```{r}
# COMPLETED CODE BELOW
census_df |>
  summarise(coldest_average_temperature_f = min(avg_temp_f))
```

To see which state has that temperature:

```{r}
census_df |>
  slice_min(avg_temp_f, n = 1, with_ties = FALSE) |>
  select(state, avg_temp_f)
```

# Extra Information and Challenges

## More Comparison Operators

The following operators each return `TRUE` or `FALSE`:

- `!=` checks inequality
- `>` checks whether the left value is greater
- `<` checks whether the left value is less
- `>=` checks whether the left value is greater than or equal
- `<=` checks whether the left value is less than or equal

```{r}
"hiking" != "skiing"
51 > 15
1000 < 1000
41 >= 41.5
50 <= 300
```

For example, find states with at least three national parks:

```{r}
census_df |>
  filter(national_parks_count >= 3)
```

Show only states that are not coastal:

```{r}
# COMPLETED CODE BELOW
census_df |>
  filter(!is_coastal)
```

Try other comparisons. For example, find states where the capybara trend score is greater than 40, or states without a major zoo.

```{r}
# Example answer
census_df |>
  filter(capybara > 40)
```

## Combining dplyr Functions

You can combine `filter()`, `select()`, `mutate()`, `group_by()`, and `summarise()` to answer more specific questions. For example, find the average pizza and hiking trend scores for states with an NBA team, grouped by personality profile.

```{r}
census_df |>
  filter(has_nba_team) |>
  group_by(personality_profile) |>
  summarise(
    average_pizza_score = mean(pizza),
    average_hiking_score = mean(hiking),
    number_of_states = n(),
    .groups = "drop"
  )
```

Among states with a Waffle House, find the state with the largest waffle trend score.

```{r}
# COMPLETED CODE BELOW
census_df |>
  filter(has_waffle_house) |>
  arrange(desc(waffles)) |>
  select(state, waffles, pancakes, donuts) |>
  head(1)
```

For coastal states, select their state name, personality profile, annual sunny days, and board-game search-interest score. Would the sunniest place also make a fun game-night stop?

```{r}
# COMPLETED CODE BELOW
census_df |>
  filter(is_coastal) |>
  select(
    state,
    personality_profile,
    annual_sunny_days,
    `board games`
  )
```

Create an `Outdoor_adventure` column that is `TRUE` when a state has at least 20 hiking-trend points and at least one national park. Then count outdoor adventure states in each personality profile.

```{r}
# COMPLETED CODE BELOW
census_df |>
  mutate(
    Outdoor_adventure = hiking >= 20 & national_parks_count >= 1
  ) |>
  group_by(personality_profile) |>
  summarise(
    outdoor_adventure_states = sum(Outdoor_adventure),
    average_hiking_score = mean(hiking),
    .groups = "drop"
  )
```
