---
title: "R Workshop: Census and Trends"
author: "BYU Library"
format: html
editor: visual
---

# Basics of R

## What is R?

R is a programming language that can act like a calculator and can also
perform more complex data-analysis tasks. For this workshop, we will focus on
three general types: *numbers*, *characters*, and *variables*.

```{r}
# Numbers
6
20.5
13 + 30

# Characters
"waffles"
"Brigham Young University"

# Variables
my_favorite_number <- 6
road_trip_snack <- "waffles"
```

R uses **functions** to manipulate and analyze data. Functions use
parentheses:

```{r}
# Prints text in quotes
print("Hello, data explorer!")

# Takes the square root of a number
sqrt(36)
```

## Key Concepts

We use the **assignment operator** (`<-` or `=`) to store information in
variables. It means "gets saved as."

```{r}
# The number 50 gets saved as the variable states_in_the_dataset
states_in_the_dataset <- 50

# Data from a CSV file gets saved as census_df
census_df <- read.csv(
  "census_and_trends_data.csv",
  check.names = FALSE
)
```

The assignment operator is different from the **comparison operator** (`==`).
A comparison returns either `TRUE` or `FALSE`.

```{r}
5 == 5
"waffles" == "tacos"
```

A **data frame** is a table or spreadsheet. Each row is an observation, and
each column is a variable. In this workshop, each row is a U.S. state, with
census-style facts and trend scores for food, sports, animals, and hobbies.

```{r}
# View the first six states
head(census_df)
```

**Libraries**, or **packages**, are toolboxes of functions written by other
people. This workshop uses the `dplyr` package for data manipulation.

```{r}
# Run this once in the R console if dplyr is not installed:
# install.packages("dplyr")

# Load dplyr in the current session
library(dplyr)
```

# Census and Trends Data Frame Analysis

## Setup and Initialization

Run these lines before starting the analysis:

- load the `dplyr` package
- read the census-and-trends CSV file
- view the table to make sure it loaded correctly

```{r}
# COMPLETED CODE BELOW
library(dplyr)

census_df <- read.csv(
  "census_and_trends_data.csv",
  check.names = FALSE
)

head(census_df) # You can also use names(census_df) or glimpse(census_df)
```

## Analyzing the Data

During this workshop, we will use these common `dplyr` functions:

- `select()` keeps chosen columns
- `filter()` keeps rows that meet a condition
- `mutate()` creates or changes columns
- `group_by()` creates groups for a later calculation
- `summarise()` calculates a summary for each group
- `glimpse()` previews a data frame's structure
- `arrange()` sorts rows

### Selecting columns

Suppose you are planning a strange-but-fun road trip. Select a few useful
columns: the state name, climate, UFO sightings, waffle interest, and hiking
interest.

```{r}
census_df %>%
  select(
    state,
    avg_temp_f,
    annual_sunny_days,
    ufo_sightings_per_100k,
    waffles,
    hiking
  )
```

### Filtering rows

To see every state that has a Waffle House:

```{r}
census_df %>%
  filter(has_waffle_house)
```

### Creating columns

The dataset includes trend scores for waffles, pancakes, and donuts. Create a
simple `Brunch_score` by adding them together.

```{r}
census_df %>%
  mutate(Brunch_score = waffles + pancakes + donuts)
```

`mutate()` displays a new data frame. To save the new column for later, assign
the result back to a variable.

```{r}
census_with_brunch <- census_df %>%
  mutate(Brunch_score = waffles + pancakes + donuts)
```

### Grouping and summarizing

The states are grouped into personality profiles. Find the average temperature,
average annual sunshine, average UFO sightings, and number of states in each
profile.

```{r}
census_df %>%
  group_by(personality_profile) %>%
  summarise(
    average_temperature_f = mean(avg_temp_f),
    average_sunny_days = mean(annual_sunny_days),
    average_ufo_sightings = mean(ufo_sightings_per_100k),
    number_of_states = n(),
    .groups = "drop"
  )
```

### Inspecting a data frame

Use `glimpse()` to see column names, data types, and a compact preview of the
data. Selecting a few columns first keeps the output easy to read.

```{r}
census_df %>%
  select(
    state,
    personality_profile,
    avg_temp_f,
    annual_sunny_days,
    ufo_sightings_per_100k
  ) %>%
  glimpse()
```

### Sorting rows

Use `arrange()` to put rows in order. Here are the ten sunniest states, from
most to least sunny:

```{r}
census_df %>%
  arrange(desc(annual_sunny_days)) %>%
  select(state, annual_sunny_days) %>%
  head(10)
```

That is enough to begin exploring the data!

## Practice Time

**Filtering:** Find states with at least 250 sunny days per year. These might
be good candidates for a sunny road trip.

```{r}
# COMPLETED CODE BELOW
census_df %>%
  filter(annual_sunny_days >= 250)
```

Find the row for Utah.

```{r}
# COMPLETED CODE BELOW
census_df %>%
  filter(state == "Utah")
```

**Mutating:** Create a `Brunch_score` column that adds the trend scores for
waffles, pancakes, and donuts.

```{r}
# COMPLETED CODE BELOW
census_df %>%
  mutate(Brunch_score = waffles + pancakes + donuts)
```

Create a Boolean `UFO_hotspot` column that is `TRUE` when a state has at least
60 UFO sightings per 100,000 people.

```{r}
# COMPLETED CODE BELOW
census_df %>%
  mutate(UFO_hotspot = ufo_sightings_per_100k >= 60)
```

**Selecting:** Create a smaller data frame with a state's personality profile,
capybara trend score, and Dungeons & Dragons trend score.

```{r}
# COMPLETED CODE BELOW
census_df %>%
  select(state, personality_profile, capybara, `dungeons and dragons`)
```

Now select only the state names.

```{r}
# COMPLETED CODE BELOW
census_df %>%
  select(state)
```

**Summarizing:** Find the total population represented in the dataset.

```{r}
# COMPLETED CODE BELOW
census_df %>%
  summarise(total_population = sum(total_pop))
```

Find the coldest average temperature in the dataset.

```{r}
# COMPLETED CODE BELOW
census_df %>%
  summarise(coldest_average_temperature_f = min(avg_temp_f))
```

To see which state has that temperature:

```{r}
census_df %>%
  slice_min(avg_temp_f, n = 1, with_ties = FALSE) %>%
  select(state, avg_temp_f)
```

# Extra Information and Challenges

## More Comparison Operators

The following operators each return `TRUE` or `FALSE`:

- `!=` checks inequality
- `>` checks whether the left value is greater
- `<` checks whether the left value is less
- `>=` checks whether the left value is greater than or equal
- `<=` checks whether the left value is less than or equal

```{r}
"hiking" != "skiing"
51 > 15
1000 < 1000
41 >= 41.5
50 <= 300
```

For example, find states with at least three national parks:

```{r}
census_df %>%
  filter(national_parks_count >= 3)
```

Show only states that are not coastal:

```{r}
# COMPLETED CODE BELOW
census_df %>%
  filter(!is_coastal)
```

Try other comparisons. For example, find states where the capybara trend score
is greater than 40, or states without a major zoo.

```{r}
# Example answer
census_df %>%
  filter(capybara > 40)
```

## Combining dplyr Functions

You can combine `filter()`, `select()`, `mutate()`, `group_by()`, and
`summarise()` to answer more specific questions. For example, find the average
pizza and hiking trend scores for states with an NBA team, grouped by
personality profile.

```{r}
census_df %>%
  filter(has_nba_team) %>%
  group_by(personality_profile) %>%
  summarise(
    average_pizza_score = mean(pizza),
    average_hiking_score = mean(hiking),
    number_of_states = n(),
    .groups = "drop"
  )
```

Among states with a Waffle House, find the state with the largest waffle trend
score.

```{r}
# COMPLETED CODE BELOW
census_df %>%
  filter(has_waffle_house) %>%
  arrange(desc(waffles)) %>%
  select(state, waffles, pancakes, donuts) %>%
  head(1)
```

For coastal states, select their state name, personality profile, annual sunny
days, and craft breweries per 100,000 people.

```{r}
# COMPLETED CODE BELOW
census_df %>%
  filter(is_coastal) %>%
  select(
    state,
    personality_profile,
    annual_sunny_days,
    craft_breweries_per_100k
  )
```

Create an `Outdoor_adventure` column that is `TRUE` when a state has at least
20 hiking-trend points and at least one national park. Then count outdoor
adventure states in each personality profile.

```{r}
# COMPLETED CODE BELOW
census_df %>%
  mutate(
    Outdoor_adventure = hiking >= 20 & national_parks_count >= 1
  ) %>%
  group_by(personality_profile) %>%
  summarise(
    outdoor_adventure_states = sum(Outdoor_adventure),
    average_hiking_score = mean(hiking),
    .groups = "drop"
  )
```
